---
title: "Exploratory Data Analysis"
output: html_notebook
---

```{r}
# Load necessary libraries
library(dplyr)

```

# Data Set Information

Basketball Reference's WNBA per-100-possession data for the seasons 1997 to 2024

| Stat   | Description                                                      |
|-----------------------------------------|-------------------------------|
| Player | Player Name                                                    |
| Team   | Team |
| G      | Games |
| MP     | Minutes Played |
| GS     | Games Started |
| FG     | Field Goals (includes both 2-point and 3-point field goals)      |
| FGA    | Field Goal Attempts (includes both 2-point and 3-point attempts) |
| FG%    | Field Goal Percentage; formula: FG / FGA.                        |
| 3P     | 3-Point Field Goals                                              |
| 3P%    | 3-Point Field Goal Percentage; the formula is 3P / 3PA.          |
| 3PA    | 3-Point Field Goal Attempts                                      |
| 2P     | 2-Point Field Goals                                              |
| 2PA    | 2-Point Field Goal Attempts                                      |
| 2P%    | 2-Point Field Goal Percentage; the formula is 2P / 2PA.          |
| FT     | Free Throws                                                      |
| FT%    | Free Throw Percentage; formula: FT / FTA.                        |
| FTA    | Free Throw Attempts                                              |
| ORB    | Offensive Rebounds                                               |
| TRB    | Total Rebounds                                                   |
| AST    | Assists                                                          |
| STL    | Steals                                                           |
| BLK    | Blocks                                                           |
| TOV    | Turnovers                                                        |
| PF     | Personal Fouls                                                   |
| PTS    | Points                                                           |
| Season | Season Start Year                                                |

```{r}
script_dir <- getwd()
data_directory <- file.path(script_dir, 'data')
file_path <- file.path(data_directory, 'per_100.csv')
df <- read.csv(file_path)

# Get the Shape
df_shape <- dim(df)
print(paste("Number of rows:", df_shape[1]))
print(paste("Number of columns:", df_shape[2]))

# List the available columns
columns <- colnames(df)
print("Columns:")
print(columns)
```
```{r}
# Check if those two columns that seem repeated are

similarity_g <- sum(df[, "G"] == df[, "G.1"])/length(df[, "G"])
similarity_mp <- sum(df[, "MP"] == df[, "MP.1"])/length(df[, "MP"])

if (similarity_g > 0.9) {
  print(paste("Columns G and G.1 are", round(similarity_g * 100, 2), "% identical"))
  df <- df[, -which(colnames(df) == "G.1")]
}

if (similarity_mp > 0.9) {
  print(paste("Columns MP and MP.1 are", round(similarity_mp * 100, 2), "% identical"))
  df <- df[, -which(colnames(df) == "MP.1")]
}

columns <- colnames(df)
print("New Columns:")
print(columns)

```



